{


\begin{agradecimientos}
  \textbf{\textsc{To the National Autonomous University of Mexico}}

  During the last 18 years, I have been part of this University, which has allowed me to know many people, from teachers to classmates and coworkers, who eventually become great friends. They all contributed to the knowledge and values that I acquired, helping in my personal and professional formation.

  \textbf{\textsc{To my family}}

  Thank you for being with me at every moment and for your support while I was pursuing my PhD. I especially want to thank my mother, Reina, who is a great source of motivation and inspiration.

  \textbf{\textsc{To the tutoring comittee and synodals}}

  I am grateful to Dr. Armando Castañeda, director of this thesis, for the support, advice, and patience provided while I was pursuing this PhD. He has been a great advisor to me for the last few years, sharing all his experience and knowledge. Also, I am grateful to Dr. Sergio Rajsbaum and Dr. Ricardo Marcelin for all their advice and the time spent in supervisor duties. Finally, I thank Dr. David Flores and Dr. Jorge Ortega for being members of the jury in my candidacy exam and the thesis defense.

  \textbf{\textsc{To my friends}}

  To Juan José Lopéz, Daniel Becerra, Olga Villagran, Eduardo López, Carlos Romero, Ricardo Rivas, and Juan Camacho, thank you for all the time you shared with me, which boosted me to be better every day.

  \textbf{\textsc{To CONAHCYT and DGAPA}}

  I am grateful to CONAHCYT for the national fellowship I received during my PhD and to DGAPA for the financial support from UNAM PAPIIT projects IN108720 and IN108723, which helped to buy the infrastructure where all the experiments of this thesis were performed.

\end{agradecimientos}
}
